Cross Validation Approaches for Penalized Cox Regression

### 2023-08-14 sub/smimr/response.tex

\item Extensions into time-varying covariates, and, preferably, joint modeling, would be useful as well. At least the first two items (more flexible survival distributions, and dependent censoring) should be explored in your work, while perhaps extensions into joint modeling to accommodate one or more longitudinal predictors could be discussed as future work in your Discussion.

\re This is an excellent point -- additional issues may certainly arise when applying cross-validation to time-varying data. Our experience with high-dimensional Cox regression has tended to involve features that were only measured once due to the cost of these experiments, but as this cost goes down, that may certainly change. We feel that addressing these issues is beyond the scope of the current manuscript, but is definitely something worth discussing. The final paragraph of our discussion now points out these issues and discusses that future work is necessary to investigate the application of the proposed approach to these settings.

\item \label{limitations} Speaking of your Discussion section, there is no mention of any limitations of the current work. Aside from the joint modeling comment above, you should also talk about violations of linearity, and of non-proportional hazards. That is, you need to provide a number of additional sentences, and references, as warranted, to talk about how the current contributions could ultimately be made more generalizable to various survival dataset characteristics, and not just those with a very simplifying structure. Some discussion on how to think about survival datasets with rare events should be commented on as well.

  \re Our point of view in this manuscript is that you have decided to fit a Cox proportional hazard model, but aren't sure what the best way to carry out cross-validation is. Certainly, if the Cox model is wrong in the first place, then it may be best to take a completely different approach to modeling, and this manuscript is not likely to be relevant. We don't really see the need to discuss issues like non-proportional hazards and non-linear effects, as they have already been covered in hundreds, if not thousands, of other manuscripts and textbooks.

@@

\item In the application section, you mention fitting elastic net models to the gene expression dataset, since elastic net does a better job than traditional Lasso when modeling many correlated genes. However, there is evidence that newer Lasso approaches (e.g., Precision Lasso: Wang et al, 2019, Bioinformatics) outperform elastic net models in this setting, so you need to justify why only the elastic net is considered as the only alternative to traditional Lasso here. Perhaps you can mention potential alternatives to elastic net as an improvement in Section 4, and include this as an area of future focus in the Discussion.

  \re Thank you for the suggestion; we now mention the precision lasso method in the discussion section.

\item Regarding elastic net, it was not justified why other alpha values (beyond 0.3, 0.5, 0.7, and 0.9) are not considered in the analysis, such as a denser grid between 0.1 and 0.5, particularly after seeing the results (in Table 3) from the sparse selection of alpha values used. More alpha values should be considered.

  \re This is certainly a fair point. In the revised manuscript, we include $\alpha=0.1$ and $\alpha=0.2$ in the comparison as well. In our experience (for all kinds of high-dimensional regression, not just survival models), there is rarely an obvious choice for $\alpha$: many choices of $\alpha$ perform similarly. For the case study presented in Section 4, it seems clear that lower values of $\alpha$ consistently outperform values near 1, but honestly, you could make a case for any value of $\alpha$ below 0.5. We focus on $\alpha=0.3$ for the sake of illustration, but the patterns (in terms of how the CV methods compare to each other) are similar for other choices of $\alpha$. In addition to modifying Table 3 itself, we have modified the text surrounding it to point out these issues, rather than the overly simplistic remark in the original submission that ``$\alpha=0.3$ is best.''

@@

\item Did the authors make the software available for the proposed approaches?

\re Yes; the proposed cross-validated linear predictor approach is implemented in the \textbf{ncvreg} package. The basic approach and the Verweij and Van Houwelingen's approach are implemented in the \textbf{glmnet} package. The code to reproduce our simulations and analyses are available as a GitHub repository.

\item Many places require major/minor clarification and corrections.

  \re Thank you very much for catching and pointing out all of these issues! We truly appreciate your attention to detail in reading our work. All of them have been fixed in the revised version of the manuscript.

### 2023-08-14 main.tex
Editing text surrounding Table 3 and response

For this analysis, we fit elastic net models to the data in addition to lasso models, as these models often improve upon the lasso when many correlated genes are involved \citep{Zou2005}. The clinical variables stage, age and gender were included in the model without penalization, as it is reasonable to assume that all three have an effect on survival. In addition to the regularization parameter $\lam$, the elastic net model has an extra tuning parameter $\alpha$, where $\alpha=1$ corresponds to lasso and $\alpha=0$ corresponds to ridge regression. For $\alpha \in \{0.3, 0.5, 0.7, 0.9, 1\}$, we selected a regularization parameter $\lam$ that has the lowest cross-validation error for each of the four cross-validation methods described in Section~\ref{Sec:methods}.  Although each method can be used to select a pair of $\alpha$ and $\lam$ values, the methods are on different scales and cannot be directly compared with each other.  For this reason, Table~\ref{tb:shedden} reports the cross-validated C-Index and Brier Scores for each method's choice of $\lam$ at each value of $\alpha$.  As shown in the table, the elastic net model with $\lam$ and $\alpha=0.3$ selected by the cross-validated linear predictor approach has the best performance in terms of C Index and Brier Score.

\begin{table}[ht]
\centering
\caption{\label{tb:shedden} The C Index and Brier Score of all elastic net models selected by cross-validation methods }
\begin{tabular}{lccccc}
\hline
  & \multicolumn{5}{c}{Cross-validated C Index} \\ 
  & $\alpha = 0.3$ & $\alpha = 0.5$ & $\alpha = 0.7$ & $\alpha = 0.9$ & $\alpha = 1$ \\ \cline{2-6}
V \& VH      & 0.639 & 0.637 & 0.638 & 0.637 & 0.637\\
Basic       & 0.647 & 0.639 & 0.638 & 0.638 & 0.639 \\
Linear Pred    & \textbf{0.650} & 0.647 & 0.638 & 0.638 & 0.639  \\
Deviance Resid & 0.645 & 0.637 & 0.638 & 0.637 & 0.637 \\ 
\addlinespace
  & \multicolumn{5}{c}{Brier Score} \\ 
  & $\alpha = 0.3$ & $\alpha = 0.5$ & $\alpha = 0.7$ & $\alpha = 0.9$ & $\alpha = 1$ \\ \cline{2-6}
V \& VH  & 0.243 & 0.244 & 0.244 & 0.245 & 0.245 \\
Basic  & 0.233 & 0.243 & 0.244 & 0.244 & 0.244 \\
Linear Pred & \textbf{0.229} & 0.232 & 0.244 & 0.244 & 0.244 \\
Deviance Resid & 0.235 & 0.244 & 0.244 & 0.245 & 0.245 \\ 
\hline
\end{tabular}
\end{table}

For each fixed $\alpha$, we compared the cross-validation errors (CVE) from the four cross-validation methods along the solution path. At $\alpha = 0.3$, CVEs are rescaled and plotted for all four methods along the same grid of $\lambda$ values in Figure~\ref{fig:CVE}. The selected $\lambda$ values, where a given CVE curve reaches its lowest point, are indicated with dotted lines.  The CVE curves for the basic and linear predictor approaches have greater curvature near their minima, and thus offer clearer support for the selected value of $\lam$.  In comparison, the deviance residual and V\&VH approaches are flatter near their minima, supporting a considerably wider range of $\lam$ values and leading to greater model uncertainty.   The selected $\lambda$ values, represented by the vertical dashed lines, and their corresponding models, reflect what was observed in the simulation studies. The linear predictor approach was most liberal and selected 85 genes. While the basic approach and deviance residual approach selected 68 genes and 60 genes respectively, Verweij and Van Houwelingen's approach was most conservative, selecting only 34 genes.  Similar patterns were observed for other $\alpha$ values.

@@

Cross-validation is a crutial step in fitting penalized Cox regression models. In this paper, we propose two cross-validation methods for Cox Regression: cross-validated linear predictors and cross-validated deviance residuals. We conducted a series of simulation experiments to study and compare their performance with two existing approaches, basic cross-validation and Verweij and Van Houwelingen's cross-validated partial likelihood, in selecting tuning parameters when fitting penalized regression models.

>>>>

For this analysis, we fit elastic net models to the data in addition to lasso models, as these models often improve upon the lasso when many correlated genes are involved \citep{Zou2005}. The clinical variables stage, age and gender were included in the model without penalization, as it is reasonable to assume that all three have an effect on survival. In addition to the regularization parameter $\lam$, the elastic net model has an extra tuning parameter $\alpha$, where $\alpha=1$ corresponds to lasso and $\alpha=0$ corresponds to ridge regression. For $\alpha \in \{0.1, 0.2, 0.3, 0.5, 0.7, 0.9, 1\}$, we selected the regularization parameter $\lam$ that has the lowest cross-validation error for each of the four cross-validation methods described in Section~\ref{Sec:methods}. As shown in Table~\ref{tb:shedden}, which reports the cross-validated C-Index and Brier Scores for each method, models with smaller values of $\alpha$ tend to perform better than models with $\alpha \approx 1$ (higher C-Index, lower Brier scores). As is typical in high-dimensional regression, there is no obvious best choice for $\alpha$, but for the sake of illustration, we focus on $\alpha=0.3$ in the remainder of this section, as it seems to consistently outperform $\alpha=1$ and there is little evidence of further gain by dropping $\alpha$ below 0.3.

\begin{table}[ht]
\centering
\caption{\label{tb:shedden} The C Index and Brier Score of all elastic net models selected by cross-validation methods }
\begin{tabular}{lccccccc}
\hline
  & \multicolumn{7}{c}{Cross-validated C Index} \\ 
  & $\alpha = 0.1$ & $\alpha = 0.2$ & $\alpha = 0.3$ & $\alpha = 0.5$ & $\alpha = 0.7$ & $\alpha = 0.9$ & $\alpha = 1$ \\ \cline{2-8}
V \& VH        & 0.645 &0.643 &0.630 &0.626 &0.617 &0.617 &0.617 \\
Basic          & 0.649 &0.650 &0.650 &0.649 &0.650 &0.651 &0.651 \\
Linear Pred    & 0.648 &0.650 &0.650 &0.649 &0.648 &0.617 &0.617 \\
Deviance Resid & 0.645 &0.647 &0.646 &0.626 &0.617 &0.617 &0.617 \\ 
\addlinespace
  & \multicolumn{7}{c}{Brier Score} \\ 
  & $\alpha = 0.1$ & $\alpha = 0.2$ & $\alpha = 0.3$ & $\alpha = 0.5$ & $\alpha = 0.7$ & $\alpha = 0.9$ & $\alpha = 1$ \\ \cline{2-8}
V \& VH        &0.231 &0.232 &0.243 &0.245 &0.249 &0.249 &0.249 \\
Basic          &0.226 &0.226 &0.225 &0.224 &0.221 &0.221 &0.221 \\
Linear Pred    &0.228 &0.226 &0.225 &0.224 &0.224 &0.249 &0.249 \\
Deviance Resid &0.231 &0.229 &0.228 &0.245 &0.249 &0.249 &0.249 \\ 
\hline
\end{tabular}
\end{table}

%% \begin{table}[ht]
%% \centering
%% \caption{\label{tb:shedden} The C Index and Brier Score of all elastic net models selected by cross-validation methods }
%% \begin{tabular}{lcccccccc}
%% \hline
%%   & \multicolumn{8}{c}{Cross-validated C Index} \\ 
%%   & $\alpha = 0.05$ & $\alpha = 0.1$ & $\alpha = 0.2$ & $\alpha = 0.3$ & $\alpha = 0.5$ & $\alpha = 0.7$ & $\alpha = 0.9$ & $\alpha = 1$ \\ \cline{2-9}
%% V \& VH     & 0.645 &0.645 &0.643 &0.630 &0.626 &0.617 &0.617 &0.617\\
%% Basic       &0.652 &0.649 &0.650 &0.650 &0.649 &0.650 &0.651 &0.651 \\
%% Linear Pred    &0.649 &0.648 &0.650 &0.650 &0.649 &0.648 &0.617 &0.617  \\
%% Deviance Resid &0.648 &0.645 &0.647 &0.646 &0.626 &0.617 &0.617 &0.617 \\ 
%% \addlinespace
%%   & \multicolumn{8}{c}{Brier Score} \\ 
%%   & $\alpha = 0.05$ & $\alpha = 0.1$ & $\alpha = 0.2$ & $\alpha = 0.3$ & $\alpha = 0.5$ & $\alpha = 0.7$ & $\alpha = 0.9$ & $\alpha = 1$ \\ \cline{2-9}
%% V \& VH &0.232 &0.231 &0.232 &0.243 &0.245 &0.249 &0.249 &0.249 \\
%% Basic  &0.221 &0.226 &0.226 &0.225 &0.224 &0.221 &0.221 &0.221 \\
%% Linear Pred & 0.220 &0.228 &0.226 &0.225 &0.224 &0.224 &0.249 &0.249 \\
%% Deviance Resid &0.229 &0.231 &0.229 &0.228 &0.245 &0.249 &0.249 &0.249 \\ 
%% \hline
%% \end{tabular}
%% \end{table}

For each fixed $\alpha$, we compared the cross-validation errors (CVE) from the four cross-validation methods along the solution path. At $\alpha = 0.3$, CVEs are normalized so that they can be compared on the same scale and plotted for all four methods along the same grid of $\lambda$ values in Figure~\ref{fig:CVE}. The selected $\lambda$ values, where a given CVE curve reaches its lowest point, are indicated with dotted lines.  The CVE curves for the basic and linear predictor approaches have greater curvature near their minima, and thus offer clearer support for the selected value of $\lam$.  In comparison, the deviance residual and V\&VH approaches are flatter near their minima, supporting a considerably wider range of $\lam$ values and leading to greater model uncertainty.   The selected $\lambda$ values, represented by the vertical dashed lines, and their corresponding models, reflect what was observed in the simulation studies. The linear predictor approach was most liberal and selected 85 genes. While the basic approach and deviance residual approach selected 68 genes and 60 genes respectively, Verweij and Van Houwelingen's approach was most conservative, selecting only 34 genes.  Similar patterns were observed for other $\alpha$ values.

@@

Cross-validation is a crucial step in fitting penalized Cox regression models. In this paper, we propose two cross-validation methods for Cox Regression: cross-validated linear predictors and cross-validated deviance residuals. We conducted a series of simulation experiments to study and compare their performance with two existing approaches, basic cross-validation and Verweij and Van Houwelingen's cross-validated partial likelihood, in selecting tuning parameters when fitting penalized regression models.
